---
title: Google Earth Engine in R
author: Abhishek Kumar
date: '2021-10-16'
slug: google-earth-engine-in-r
categories:
  - R
  - rgee
tags:
  - Remote Sensing
  - Google Earth Engine
subtitle: ''
summary: ''
authors: []
lastmod: '2021-10-16T16:23:06+05:30'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []

bibliography: [refs.bib, packages.bib]
---

```{r message=FALSE, warning=FALSE, echo=FALSE}
## set global options for markdown
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, fig.align = 'center',
			    fig.width = 7, fig.height = 5, out.width = '90%', dev = "svglite",
			    echo=T, message=FALSE, warning=FALSE)
```


## Introduction

Google Earth Engine (GEE) is a cloud-based platform for planetary-scale geo-spatial analysis [@Gorelick2017]. The cloud-based computation ability overcomes the limitations of data access, data storage, personal limited computation ability and time. Thus, it provides a platform for large scale analysis of earth observations through access to satellite imagery.

Google offered support only for Python and JavaScript, and GEE was not accessible to `R` [@R-base] users until Aybar et al. [-@rgee2020] developed `rgee` package [@R-rgee].

```{r message=FALSE, warning=FALSE, results='hide'}
library(rgee)
ee_Initialize()

#knitr::write_bib(x = c(.packages(), "blogdown"), "packages.bib")
```

## Data and Pre-processing

Data Product: [Sentinel-2 MSI: MultiSpectral Instrument, Level-2A](https://developers.google.com/earth-engine/datasets/catalog/COPERNICUS_S2_SR)

Sentinel-2 is a wide-swath, high-resolution, multi-spectral imaging mission supporting Copernicus Land Monitoring studies, including the monitoring of vegetation, soil and water cover, as well as observation of inland waterways and coastal areas. The SENTINEL-2 data contain 13 spectral bands representing TOA reflectance scaled by 10000.

```{r}
# define a region of interest
roi <- ee$Geometry$Rectangle(76.85, 30.66, 77.01, 30.75)

# load and pre-process sentinel-2 data
# Filter the image collection
img <- ee$ImageCollection("COPERNICUS/S2_SR")$
  
  # filter the date range we are interested in
  filterDate("2020-01-01", "2020-12-31")$
  
  # filter geographic area we are interested in 
  filterBounds(roi)$
  
  # filter cloud cover less than 20%
  filter(ee$Filter$lt("CLOUD_COVERAGE_ASSESSMENT", 20))$
  
  # filter cloud pixel percentage to less than 20%
  filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 20))$
  
  # reduce image collection by taking median value
  median()$
  
  # clip to region of interest
  clip(roi)

# print band names
img$bandNames()$getInfo()
```

## Median true color composite

```{r}
Map$centerObject(roi, 12)
Map$addLayer(img, 
             visParams = list(min = 0, max = 3000, bands = c("B4", "B3", "B2")),
             name = "true color")
```

## Median false color composite

```{r}
Map$centerObject(roi, 12)
Map$addLayer(img, 
             visParams = list(min = 0, max = 3000, bands = c("B8", "B4", "B3")),
             name = "false color")
```

## Spectral Indices

### Vegetation cover

#### NDVI: Normalized Difference Vegetation Index

```{r}
# calculate indices e.g. NDVI
ndvi <- img$expression(
  expression = "(NIR - RED)/(NIR + RED)",
  opt_map = list("RED" = img$select("B4"),
                 "NIR" = img$select("B8")
                 ))$rename("ndvi")

vis_pal <- c('FFFFFF', 'CE7E45', 'DF923D', 'F1B555', 'FCD163', '99B718', 
             '74A901', '66A000', '529400', '3E8601', '207401', '056201',
             '004C00', '023B01', '012E01', '011D01', '011301')

Map$centerObject(roi, 12)
Map$addLayer(ndvi, 
             visParams = list(min = 0, max = 1, palette = vis_pal),
             name = "NDVI")
```

#### NDVIre: Normalized Difference Vegetation Index red-edge

```{r}
ndvire <- img$expression(
  expression = "(RedEdge1 - RED)/(RedEdge1 + RED)",
  opt_map = list("RED" = img$select("B4"),
                 "RedEdge1" = img$select("B5")
                 ))$rename("ndvire")

vis_pal <- c('FFFFFF', 'CE7E45', 'DF923D', 'F1B555', 'FCD163', '99B718', 
             '74A901', '66A000', '529400', '3E8601', '207401', '056201',
             '004C00', '023B01', '012E01', '011D01', '011301')

Map$centerObject(roi, 12)
Map$addLayer(ndvire, 
             visParams = list(min = 0, max = 1, palette = vis_pal),
             name = "NDVIre")
```

#### EVI: Enhanced vegetation Index

```{r}
# calculate EVI
evi <- img$expression(
  expression = "2.5 * ((NIR - RED) / (NIR + 6 * RED - 7.5 * BLUE + 10000))",
  opt_map = list("BLUE" = img$select("B2"),
                 "RED" = img$select("B4"),
                 "NIR" = img$select("B8")
                 ))$rename("evi")

Map$centerObject(roi, 12)
Map$addLayer(evi, 
             visParams = list(palette = vis_pal),
             name = "EVI")
```

#### SAVI: Soil-adjusted Vegetation Index

```{r}
savi <- img$expression(
  expression = "((NIR - RED) / (NIR + RED + 0.5)) * 1.5",
  opt_map = list("RED" = img$select("B4")$multiply(0.0001),
                 "NIR" = img$select("B8")$multiply(0.0001)
                 ))$rename("savi")

Map$centerObject(roi, 12)
Map$addLayer(savi, 
             visParams = list(min = 0, max = 0.5, 
                              palette = c("brown", "yellow", "green")),
             name = "SAVI")
```

### Water bodies

#### NBR/NDWI: Normalized Burn Ratio / Normalized Difference Water Index

```{r}
ndwi <- img$expression(
  expression = "(NIR - SWIR) / (NIR + SWIR)",
  opt_map = list("NIR" = img$select("B8"),
                 "SWIR" = img$select("B12")
                 ))$rename("ndwi")

Map$centerObject(roi, 12)
Map$addLayer(ndwi, 
             visParams = list(min = 0, max = 0.5, 
                              palette = c("blue", "white", "green")),
             name = "NDWI")
```

#### MNDWI: Modified Normalized Difference Water Index

```{r}
mndwi <- img$expression(
  expression = "(GREEN - SWIR1) / (GREEN + SWIR1)",
  opt_map = list("GREEN" = img$select("B3"),
                 "SWIR1" = img$select("B11")
                 ))$rename("mndwi")

Map$centerObject(roi, 12)
Map$addLayer(ndwi, 
             visParams = list(min = 0, max = 0.5, 
                              palette = c("blue", "white", "green")),
             name = "MNDWI")
```

#### AWEI_sh: Automated Water Extraction Index - shadow

```{r}
aweish <- img$expression(
  expression = "BLUE + 2.5 * GREEN - 1.5 * (NIR + SWIR1) - 0.25 * SWIR2",
  opt_map = list("BLUE" = img$select("B2"),
                 "GREEN" = img$select("B3"),
                 "NIR" = img$select("B8"),
                 "SWIR1" = img$select("B11"),
                 "SWIR2" = img$select("B12")
                 ))$rename("aweish")

Map$centerObject(roi, 12)
Map$addLayer(aweish, 
             visParams = list(min = 0, max = 1, palette = c("white", "blue")),
             name = "AWEIsh")
```

### Built up areas

#### NDBI: Normalized Difference Built-up Index

```{r}
ndbi <- img$expression(
  expression = "(SWIR1 - NIR) / (SWIR1 + NIR)",
  opt_map = list("NIR" = img$select("B8"),
                 "SWIR1" = img$select("B11")
                 ))$rename("ndbi")

Map$centerObject(roi, 12)
Map$addLayer(ndbi, 
             visParams = list(min = -1, max = 1, palette = c("white", "brown")),
             name = "NDBI")
```

#### BUI: Built-up Index

```{r}
bui <- img$expression(
  expression = "((SWIR1 - NIR) / (SWIR1 + NIR)) - ((NIR - RED)/(NIR + RED))",
  opt_map = list("RED" = img$select("B4"),
                 "NIR" = img$select("B8"),
                 "SWIR1" = img$select("B11")
                 ))$rename("bui")


Map$centerObject(roi, 12)
Map$addLayer(bui, 
             visParams = list(min = -1, max = 0.5, palette = c("green", "white", "brown")),
             name = "BUI")
```

### Soil related Indices

#### NDTI: Normalized Difference Tillage Index

```{r}
ndti <- img$expression(
  expression = "(SWIR1 - SWIR2) / (SWIR1 + SWIR2)",
  opt_map = list("SWIR1" = img$select("B11"),
                 "SWIR2" = img$select("B12")
                 ))$rename("ndti")

Map$centerObject(roi, 12)
Map$addLayer(ndti, 
             visParams = list(min = 0, max = 0.5, palette = c("brown", "yellow", "green")),
             name = "NDTI")
```

#### BSI: Bare Soil Index

```{r}
bsi <- img$expression(
  expression = "(((SWIR1 + RED) - (NIR + BLUE)) / (SWIR1 + RED)) + (NIR + BLUE)",
  opt_map = list("BLUE" = img$select("B2"),
                 "RED" = img$select("B4"),
                 "NIR" = img$select("B8"),
                 "SWIR1" = img$select("B11"),
                 "SWIR2" = img$select("B12")
                 ))$rename("bsi")

Map$centerObject(roi, 12)
Map$addLayer(bsi, 
             visParams = list(min = 1000, max = 4000, palette = c("white", "brown")),
             name = "BSI")
```


## References